---
title: "Data visualization with ggplot2"
author: "Joshua Cheung (PID: A15441585)"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1: Overview

No questions or code given for this section

# Section 2: Background

**Q. For which phases is data visualization important in our scientific workflows?**

All of the above

**Q.True or False? The ggplot2 package comes already installed with R?**

False

# Section 3: Getting Organized

No questions or code given for this section

# Section 4: Common Plot Types

**Q. Which plot types are typically NOT used to compare distributions of numeric variables?**

Network graphs

**Q. Which statement about data visualization with ggplot2 is incorrect?**

ggplot2 is the only way to create plots in R

# Section 5: Creating Scatter Plots

## Specifying a dataset with ggplot()

```{r}
# install.packages("ggplot2") ## un-comment if necessary
library(ggplot2)
ggplot(cars)
```

## Specifying aesthetic mappings with aes()

Now set the x and y axis coordinates

```{r}
ggplot(cars) +
  aes(x=speed, y=dist)
```

## Specifying a geom layer with geom_point()

Add geom_point() to get a point layer

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```


**Q. Which geometric layer should be used to create scatter plots in ggplot2?**

geom_point()

**Q. In your own RStudio can you add a trend line layer to help show the relationship between the plot variables with the geom_smooth() function?**

```{r}
library(ggplot2)
p <- ggplot(data=cars) +
  aes(x=speed, y=dist) +
  geom_point()

p + geom_smooth()
```

**Q. Can you finish this plot by adding various label annotations with the labs() function and changing the plot look to a more conservative "black & white" theme by adding the theme_bw() function:**

```{r}
p + labs(title = "Speed and Stopping distances of Cars",
         x = "speed (Miles per Hour)",
         y = "Stopping Distance (feet)",
         caption = "Dataset Source: 'cars'") +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
```

## Adding more plot aesthetics through aes()

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

**Q. Use the nrow() function to find out how many genes are in this dataset. What is your answer?**

There are 5196 genes in this dataset.

```{r}
nrow(genes)
```
**Q. Use the colnames() function and the ncol() function on the genes data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?**

There are 4 columns.

```{r}
colnames(genes)
ncol(genes)
```
**Q. Use the table() function on the State column of this data.frame to find out how many 'up' regulated genes there are. What is your answer?**

There are 127 up-regulated genes.

```{r}
table(genes$State)
```
**Q. Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?**

The fraction of the total genes that are up-regulated in this data set rounded at two significant figures is 2.44

```{r}
q <- table(genes$State)/nrow(genes)*100
round(q, 2)
```

**Q. Complete the code below to produce the following plot**

 The completed code is:

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2) +
  geom_point()
```

Map the state column to point color

```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
```

Change the colors to blue, gray, red.

```{r}
p + scale_colour_manual( values=c("blue","gray","red") )
```

**Q. Nice, now add some plot annotations to the p object with the labs() function so your plot looks like the following:**

```{r}
r <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, colour=State) +
  geom_point()
r + scale_colour_manual(values = c("blue", "gray", "red")) +
  labs(title = "Gene Expression Changes Upon Drug Treatment",
       x = "Control (no drug)",
       y = "Drug Treatment")
```

# Section 6 OPTIONAL: Going Further

```{r}
# install.packages("gapminder")  ## un-comment to install if necessary
# install.packages("dplyr")  ## un-comment to install if necessary
library(gapminder)
library(dplyr)
gapminder_2007 <- gapminder%>%filter(year==2007)
```

**Let's consider the gapminder_2007 dataset which contains the variables GDP per capita gdpPercap and life expectancy lifeExp for 142 countries in the year 2007**

**Q. Complete the code below to produce a first basic scatter plot of this gapminder_2007 dataset:**

The completed code is:

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point() 
```

Now make the points slightly transparent (set alpha=0.5)

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point(alpha=0.5)
```

## Adding more variables to aes()

Now map the continent variable to the point color aesthetic and population through the point size argument of aes().

```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```

Now change the plot so that the points are colored by the numeric variable: population.

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)
```

## Adjusting point size

Now change the plot so that point size is based on the population of each country.

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```

Improve the plot to reflect the actual population differences by the point size.  Use scale_size_area() instead.

```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

**Q. Can you adapt the code you have leaqrned thus far to reproduce our gapminder scatter plot for the year 1957? What do you notice abouyt this plot is it easy to compare with the one for 2007?**

**Steps to produce your 1957 plot should include**

**Use dplyr to filter the gapmider dataset to include only the year 1957 (check above for how we did this for 2007).**

**Save your result as gapminder_1957.**

**Use the ggplot() function and specify the gapminder_1957 dataset as input**

**Add a geom_point() layer to the plot and create a scatter plot showing the GDP per capita gdpPercap on the x-axis and the life expectancy lifeExp on the y-axis**

**Use the color aesthetic to indicate each continent by a different color**

**Use the size aesthetic to adjust the point size by the population pop**

**Use scale_size_area() so that the point sizes reflect the actual population differences and set the max_size of each point to 15 -Set the opacity/transparency of each point to 70% using the alpha=0.7 parameter**

```{r}
gapminder_1957 <- gapminder%>%filter(year==1957)
ggplot(gapminder_1957) +
  aes(x=gdpPercap, y=lifeExp, colour=continent, size=pop) +
  scale_size_area(max_size = 15) +
  geom_point(alpha=0.7) 
```

**Q. Do the same steps above but include 1957 and 2007 in your input dataset for ggplot(). You should now include the layer facet_wrap(~year) to produce the following plot:**

```{r}
gapminder_1957_2007 <- gapminder%>%filter(year==1957 | year==2007)
ggplot(gapminder_1957_2007) +
  aes(x=gdpPercap, y=lifeExp, colour=continent, size=pop) +
  scale_size_area(max_size = 15) +
  geom_point(alpha=0.7) +
  facet_wrap(~year)
```

# Section 7 OPTIONAL: Bar Charts

## Introduction to bar charts

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

## Creating a simple bar chart

Now create a bar chart with gapminder_top5 dataset (countries are in alphabetical order by default).

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```

**Q. Plot life expectancy by country**

**Create a bar chart showing the life expectancy of the five biggest countries by population in 2007.**

**Use the ggplot() function and specify the gapminder_top5 dataset as input**

**Add a geom_col() layer to the plot**

**Plot one bar for each country (x aesthetic)**

**Use life expectancy lifeExp as bar height (y aesthetic)**

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = lifeExp))
```

## Filling bars with color

Plot the population of the biggest countires and use the contient variable to color each bar.

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = continent))
```

Now use the life expectancy variable to fill the colors for the bars.

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop, fill = lifeExp))
```

**Q. Plot population size by country**

**Create a bar chart showing the population (in millions) of the five biggest countries by population in 2007.**

**Use the ggplot() function and specify the gapminder_top5 dataset as input**

**Add a geom_col() layer to the plot**

**Plot one bar for each country (x aesthetic)**

**Use population pop as bar height (y aesthetic)**

**Use the GDP per capita gdpPercap as fill aesthetic**

```{r}
ggplot(gapminder_top5) +
  geom_col(aes(x=country, y=pop, fill=gdpPercap))
```

Now change the orders of the bars

```{r}
ggplot(gapminder_top5) +
  geom_col(aes(x=reorder(country, -pop), y=pop, fill=gdpPercap))
```

Now just fill by country

```{r}
ggplot(gapminder_top5) +
  aes(x=reorder(country, -pop), y=pop, fill=country) +
  geom_col(col="gray30") +
  guides(fill=FALSE)
```

## Flipping bar charts

```{r}
head(USArrests)
```

```{r}
USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_col() +
  coord_flip()
```

A bit crowded, use an alternative custom visualization.  Combine geom_point() and geom_segment() to do this

```{r}
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_point() +
  geom_segment(aes(x=State, 
                   xend=State, 
                   y=0, 
                   yend=Murder), color="blue") +
  coord_flip()
```

# SECTION 8 ADVANCED: Plot Animation

```{r}
# install.packages("gifski") ## un-comment to install if necessary
# install.packages("gganimate") ## un-comment to install if necessary
```

For this exercise, we only show the images first and last year generated for this pdf (1952 and 2007 respectively).

```{r}
library(gganimate)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  # Continent facet wrap
  facet_wrap(~continent) +
  # The following code is for gganimate
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

# Section 9: Combining Plots

```{r}
#install.packages("patchwork") ## un-comment to install if needed
```


```{r}
library(patchwork)

# These are example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# This part is for pathcwork
(p1 | p2 | p3) /p4
```

